Stellar Mass and Stellar Kinematics Can Constrain Gravitational Lens Models

### 2014-07-20 limits/ms.tex

\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$45 avenue des Etats-Unis, 78035, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Physik-Institut, Universit\"at Z\"urich, 190 Winterthurerstrasse, 8057, Z\"urich, Switzerland
}

@@

In this paper we demonstrate which combination of lensing, stellar mass, and
stellar kinematic constraints best constrain the projected mass profile and
shape of several gravitational lens produced from mock data. In particular,
we introduce the idea of using estimates of the stellar mas content to place
lower bounds on the lensing mass.
We also introduce a new lens modeling framework  -- \Glass\ (Gravitational
Lensing AnalysiS Software). The compute-intensive core of \Glass\ is a parallel
implementation of the high-dimensional sampling algorithm from
\cite{2012MNRAS.425.3077L}.  Around that core, \Glass\ has extensible code (in
software terminology, a framework) in Python, encoding the lens equation, a
basis and priors for the mass distribution, and various post-processing and
visualisation modules.

@@

Unfortunately, the lens equation and the arrival times alone are not enough to
form a closed volume in the solution space. We have not imposed any strong
constraints on the density even so much as to ensure that it is non-negative
everywhere. We must therefore provide additional linear constraints so that we
only sample physical models. Below we list typical constraints that \Glass\
provides. The explicit implementation for these priors has
been explained in \cite{2008ApJ...679...17C}, but can be summarised as
%
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The azimuthally averaged density profile must have a slope everywhere $\le 0$.
\item Image parity is enforced.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item The density is inversion symmetric.
\end{enumerate}

The first three of these priors are very important for creating physically
meaningful lenses. Prior (ii) is necessary to prevent rings from forming where
the inner region is less dense. This would be unusual for a galaxy or a
cluster. Prior (iv) can be relaxed somewhat by increasing the angle. The
smoothness prior (v) requires pixel values to be less than a constant factor of
the average neighbor value. This can be relaxed by increasing the constant or
completely disabled. For this paper, we have chosen \emph{not} to use this
prior, and therefore allow for the presence of small density substructure.
Prior (vi) enforces a radial symmetry that requires radially opposed pixels
(i.e., on opposite sides of the center) to be equal. This prior is particularly
important for modelling doubles which must be symmetric. If a lens is suspected
to be symmetric this can also significantly improve the modelling time for
quads.

@@

\begin{figure*}
\includegraphics[width=0.85\textwidth]{BCarrival_surfaces}
\caption{The lens configurations for the six test cases using the \mockBC{}
mock galaxy. The other mock galaxies produce similar results. Here, the central
image is shown, although not all tests include it. The naming convention
indicates the redshift of the sources with Z1=1.72, Z2=0.72, and Z3=0.51.  The
central image only belongs to the Z1 source to avoid overconstraining the
models. Image colors represent a common source. Grey circles are a visual aid
to help determine radial separation between images. The axes are in
arcseconds.}
\label{arrival surfaces}
\end{figure*}

@@

\figref{main results} and \figref{main results pixel-wise} show the results for
our full mock data ensemble. Each subplot corresponds to a different mock
galaxy, as marked. We show the fractional error of the mass distribution for
each of the test configurations with (red) and without (black) stellar mass. In
\figref{main results} we define the error
%
\begin{equation} \label{ferror R}
  f_R = \frac {\sum_i \left|M(i) - \tilde M(i)\right| } {\sum \tilde M(i)}
\end{equation}
%
based on the mass $M(i)$ of each pixel ring $i$ and the mass $\tilde M$ from the mock galaxy. 
In \figref{main results pixel-wise} the error is defined over all the pixels $\vec\theta$:
%
\begin{equation} \label{ferror theta}
f_\theta = \frac {\sum_{\vec\theta} \left|M(\vec\theta) - \tilde M(\vec\theta)\right| } {\sum_{\vec\theta} \tilde M(\vec\theta)}
\end{equation}
%
Since both error measurements consider the mass of each pixel, we are
implicitly weighting the recovered density by the varying size of the pixels.
The value $f_R$ emphasizes the error one would see from radial profiles, while
$f_\theta$ is useful as a measure of how well each individual pixel is recovered.
For both $f_R$ and $f_\theta$, we only consider mass up to one pixel length passed the
outermost image, since there is no longer any lensing information beyond that
point. This means we typically use 8 bins, linearly spaced, ignoring the
outermost 3 bins.  The spacing changes, however, at the border between the high
resolution region in the middle.

@@

We can also more directly probe the recovery of the shape of the mass
distribution by considering the ratio of the major and minor axes $\lambda_1,
\lambda_2$ of the inertia ellipse. If they are equal, the mass is distributed
uniformly on the projected disc. The more dissimilar they are, the more
elliptical the mass distribution. We define the global measure of lens shape as
%
\begin{equation} 
    s \equiv \lambda_1/\lambda_2
\end{equation} 
%
where $\lambda_1$ and $\lambda_2$ are the eigenvalues of the 2D inertia tensor:
%
\begin{equation}
\left(
\begin{matrix}
   \sum_{\vec\theta} M(\vec\theta) \theta^2_y 
    & - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y \\
    - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y
    & \sum_{\vec\theta} M(\vec\theta) \theta^2_x
\end{matrix}
\right)
    \label{eqn:inertia}
\end{equation}
%
We always take $\lambda_1$ to be the largest value. As with $f_R$ and
$f_\theta$, we only consider mass up to one radial position passed the
outermost image and compute the fractional error as:
%
\begin{equation} \label{ferror shape}
  f_\mathrm{shape} = \left|s - \tilde s\right|  / \tilde s
\end{equation}
%
where $\tilde s$ is the shape of the mock galaxy. The distribution of
$f_\mathrm{shape}$ for each mock galaxy and each test case is shown in
\figref{shape results}.  Interestingly, for this global shape parameter
recovery it appears more important to have time delay data and/or a central
image (TD,TD+central) than to have a quad or multiple sources with wide redshift
separation. In all cases, the stellar mass little-aids the recovery, reflecting
the fact that $s$ is heavily weighted towards the shape at the {\it edge} of
the mass map, rather than at the centre where the stars may dominate the
potential (see \eqnref{eqn:inertia}).

>>>>

%\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\title[Lens Recovery with \Glass]{Stellar Mass and Stellar Kinematics Can Constrain Gravitational Lens Models}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$45 avenue des Etats-Unis, 78035, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Physik-Institut, Universit\"at Z\"urich, 190 Winterthurerstrasse, 8057, Z\"urich, Switzerland
}

@@

In this paper we demonstrate which combination of lensing, stellar mass, and
stellar kinematic constraints best constrain the projected mass profile and
shape of several gravitational lens produced from mock data. In particular,
we introduce the idea of using estimates of the stellar mas content to place
lower bounds on the lensing mass.
We have also developed a new lens modeling framework  -- \Glass\ (Gravitational
Lensing AnalysiS Software) which improves upon the earlier techniques. The
key developments are

\begin{enumerate}
\item A modular framework that allows new priors to be added and modified easily.
\item The basis functions approximating a model can be changed.
\item Non-linear constraints can be imposed in an automated post-processing step.
\item The central region can have a higher resolution to capture steep models.
\item Stellar density can be used as an additional constraint.
\item Point or extended mass objects can be placed in the field.
\item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}

The compute-intensive core of \Glass\ is a parallel
implementation of the high-dimensional sampling algorithm from
\cite{2012MNRAS.425.3077L}.  Around that core, \Glass\ has extensible code (in
software terminology, a framework) in Python, encoding the lens equation, a
basis and priors for the mass distribution, and various post-processing and
visualisation modules.

@@

The lens equation and the arrival times alone are not enough to form a closed
volume in the solution space. The data alone do not even constraint the density
to be non-negative everywhere. We therefore provide additional linear
constraints so that only physical models are sampled. 
\Glass\ provides a number of typical constraints similar to those discussed in
\cite{2008ApJ...679...17C}: 
%
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The azimuthally averaged density profile must have a slope everywhere $\le 0$.
\item Image parity is enforced.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item The density is inversion symmetric.
\end{enumerate}

The first three of these priors are very important for creating physically
meaningful lenses. Prior (ii) is necessary to prevent rings from forming where
the inner region is less dense. This would be unusual for a galaxy or a
cluster. Prior (iv) can be relaxed somewhat by increasing the angle. The
smoothness prior (v) requires pixel values to be less than a constant factor
(normally 2) of the average neighbor value. This can be relaxed by increasing
the constant or completely disabled. For this paper, we have chosen \emph{not}
to use this prior, and therefore allow for the presence of small density
substructure.  Prior (vi) enforces a radial symmetry that requires radially
opposed pixels (i.e., on opposite sides of the center) to be equal. This prior
is particularly important for modelling doubles which must be symmetric. If a
lens is suspected to be symmetric this can also significantly improve the
modelling time for quads.

@@

\begin{figure*}
\includegraphics[width=0.85\textwidth]{BCarrival_surfaces}
\caption{The lens configurations for the six test cases using the \mockBC{}
mock galaxy. The other mock galaxies produce similar results. Here, the central
image is shown, although not all tests include it. The naming convention
indicates the redshift of the sources with Z1=1.72, Z2=0.72, and Z3=0.51.  The
central image only belongs to the Z1 source to avoid overconstraining the
models. Images of the same color share a common source. The extended source examples have been
contructed so that the images will form arclets.
Grey circles are a visual aid
to help determine radial separation between images. The axes are in
arcseconds.}
\label{arrival surfaces}
\end{figure*}

@@

\figref{main results} and \figref{main results pixel-wise} show the results for
our full mock data ensemble. Each subplot corresponds to a different mock
galaxy, as marked. We show the fractional error of the mass distribution for
each of the test configurations with (red) and without (black) stellar mass. In
\figref{main results} we define the error
%
\begin{equation} \label{ferror R}
  f_R = \frac {\sum_i \left|M(i) - \widehat M(i)\right| } {\sum \widehat M(i)}
\end{equation}
%
based on the mass $M(i)$ of each pixel ring $i$ and the mass $\widehat M$ from the mock galaxy. 
In \figref{main results pixel-wise} the error is defined over all the pixels $\vec\theta$:
%
\begin{equation} \label{ferror theta}
f_\theta = \frac {\sum_{\vec\theta} \left|M(\vec\theta) - \widehat M(\vec\theta)\right| } {\sum_{\vec\theta} \widehat M(\vec\theta)}
\end{equation}
%
Since both error measurements consider the mass of each pixel, we are
implicitly weighting the recovered density by the varying size of the pixels.
The value $f_R$ emphasizes the error one would see from radial profiles, while
$f_\theta$ is useful as a measure of how well each individual pixel is recovered.
For both $f_R$ and $f_\theta$, we only consider mass up to one pixel length passed the
outermost image, since there is no longer any lensing information beyond that
point. This means we typically use 8 bins, linearly spaced, ignoring the
outermost 3 bins.  The spacing changes, however, at the border between the high
resolution region in the middle.

@@

We can also more directly probe the recovery of the shape of the mass
distribution by considering the ratio of the major and minor axes $\lambda_1,
\lambda_2$ of the inertia ellipse. If they are equal, the mass is distributed
uniformly on the projected disc. The more dissimilar they are, the more
elliptical the mass distribution. We define the global measure of lens shape as
%
\begin{equation} 
    s \equiv \lambda_1/\lambda_2
\end{equation} 
%
where $\lambda_1$ and $\lambda_2$ are the eigenvalues of the 2D inertia tensor:
%
\begin{equation}
\left(
\begin{matrix}
   \sum_{\vec\theta} M(\vec\theta) \theta^2_y 
    & - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y \\
    - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y
    & \sum_{\vec\theta} M(\vec\theta) \theta^2_x
\end{matrix}
\right)
    \label{eqn:inertia}
\end{equation}
%
We always take $\lambda_1$ to be the largest value. As with $f_R$ and
$f_\theta$, we only consider mass up to one radial position passed the
outermost image and compute the fractional error as:
%
\begin{equation} \label{ferror shape}
  f_\mathrm{shape} = \left|s - \widehat s\right|  / \widehat s
\end{equation}
%
where $\widehat s$ is the shape of the mock galaxy. The distribution of
$f_\mathrm{shape}$ for each mock galaxy and each test case is shown in
\figref{shape results}.  Interestingly, for this global shape parameter
recovery it appears more important to have time delay data and/or a central
image (TD,TD+central) than to have a quad or multiple sources with wide redshift
separation. In all cases, the stellar mass little-aids the recovery, reflecting
the fact that $s$ is heavily weighted towards the shape at the {\it edge} of
the mass map, rather than at the centre where the stars may dominate the
potential (see \eqnref{eqn:inertia}).
